On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-17 paper/application.tex
reg update

\section{Application to protoplanetary disks}\label{2dppd}
We now apply our linear framework to assess the stability of realistic 
protoplanetary disks. We consider the gravito-turbulent disk models recently
developed by \cite{rafikov15}. This 2D, Keplerian disk orbits a Solar 
mass star and is defined by the following parameters,

\begin{itemize}
  \item $\dot{M}$, the global radial mass accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. \citep[See][for details.]{rafikov15} This gives the $Q(R)$
value required for input into our linear framework. We derive other
dimensionless parameters below.

@@

\subsection{PPD cooling}
Energy loss in \citeauthor{rafikov15}'s model is given by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. Eq. \ref{ftau} accounts for cooling in the
optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
regimes.

@@

PPDs become optically-thin ($\tau\to 0$) with 
increasing distance from its host star. Then $\theta$ decreases, and
the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
5/6$ in the (gravito-turbulent) outer disk, which is always satisfied.  
On the other hand, numerical simulations of gravito-turbulence
show that the second inequality is generally satisfied \citep{rice11}.  
Taken together, this means that the outer regions of a realistic PPD,
cooling itself cannot cause instability.

%conditions are likely satisfied in the outer disk 
%Condition \ref{ppd_invisc_cond} must be violated in order for cooling
%to destablize

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied
everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation.

@@

The most unstable wavelength is a few times the disk thickness. The
increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
$\sim 100\mathrm{AU}$ occurs as the disk transitions from optically
thick to optically thin regimes. The mismatch at large distances is
not surprising since the above expressions assume $|K|\ll
1$. Nevertheless the analytic estimates reproduce the correct
behavior.

@@

%{\bf despite apply beta cooling results to realistic cooling disks}
%Our result suggest an alternative interpretation of disk fragmentation
%is where the gravito-turbulent state becomes dynamically unstable due to
%both viscosity and self-gravity.

%{\bf note: disk might be stablized by extremely large alpha at large
%  distances but `turbulent' framework no longer valid }

@@

\subsection{3D PPD with radiative diffusion}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
wavenumber for $R\in[10,100]$AU. We also plot the corresponding 2D
results with softened self-gravity. There is a close match for
$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
1$) and both models apply. However, beyond $60$AU where the disk is
optically thin, radiative diffusion (the 3D curve) does not apply and
under-estimates the growth rates. Nevertheless, the comparison shows
that the transition radius $R\sim60$AU can be calculated within the 2D
framework. This is not surprising since the instability is 
fundamentally driven by viscous and self-gravity forces in the
momentum equations, not the energy equation.

%the instability is due to viscosity/gravity in momenta equation so
%details of energy eq doesn't matter

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous, 
    self-gravitating unstable modes. The black curves are obtained
    from the 3D eigenvalue problem with radiative diffusion, while the
    green curves are 
    \label{rafikov_growth3d}}
\end{figure}

%3d rad diffusion is mathematically different problem (more
%derivatives) so worth considering briefly
%optically thin cooling in 3D is the same problem as before

%use 2D model as midplane values to get 3D structure assuming radiative
%diffusion and constant alpha w.r.t. height. allow cooling time to vary
%with height. can't assume polytropic vertical structure. need to
%reformulate linear framework (some transformations needed). 
%skip 2D eigen prob if 3D works out

>>>>

\section{Application to protoplanetary disks}\label{2dppd}
We now apply our linear framework to assess the stability of realistic 
protoplanetary disks. We consider the gravito-turbulent disk models recently
developed by \cite[][hereafter \citetalias{rafikov15}]{rafikov15}. 
This 2D, Keplerian disk orbits a Solar 
mass star and is defined by the following parameters,

\begin{itemize}
  \item $\dot{M}$, the global radial mass accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. \citepalias[See][for details.]{rafikov15} This gives the $Q(R)$
value required for input into our linear framework. We derive other
dimensionless parameters below.

@@

\subsection{PPD cooling}
Energy loss in \citetalias{rafikov15}'s model is given by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. Eq. \ref{ftau} accounts for cooling in the
optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
regimes.

@@

PPDs become optically-thin ($\tau\to 0$) with 
increasing distance from the star. Then $\theta$ decreases, and
the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
5/6$ in the outer disk, which is always satisfied.  
On the other hand, numerical simulations of gravito-turbulence
show that the second inequality is generally satisfied \citep{rice11}.  
Taken together, this means that in the outer regions of a realistic PPD,
cooling itself cannot cause instability.

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
We adopt the opacity scale $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$  as in
\citetalias{rafikov15}. We use 
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied
everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
These are set as input to the 2D dispersion relation
(Eq. \ref{thindisk}).

@@

The most unstable wavelength is a few times the disk thickness. The
increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
$\sim 100\mathrm{AU}$ occurs as the disk transitions from
optically-thick to optically-thin regimes. The mismatch at large
distances is  
not surprising since the above expressions assume $|K|\ll
1$. Nevertheless the analytic estimates reproduce the correct
behavior.

@@

%{\bf despite apply beta cooling results to realistic cooling disks}
%Our result suggest an alternative interpretation of disk fragmentation
%is where the gravito-turbulent state becomes dynamically unstable due to
%both viscosity and self-gravity. 
%{\bf note: disk might be stablized by extremely large alpha at large
%  distances but `turbulent' framework no longer valid }

@@

\subsection{3D PPD with radiative diffusion}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?} 
We briefly consider 3D PPDs with explicit radiative diffusion
(\S\ref{rad_cool}). 
Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
wavenumber for $R\in[10,100]$AU; along with corresponding 2D 
results with softened self-gravity. There is a close match for
$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
1$) and both models apply. However, beyond $60$AU where the disk
becomes optically-thin, radiative diffusion (the 3D curve) is not
valid and under-estimates the growth rates. Nevertheless, the
comparison shows that the transition radius $R_t\sim60$AU can be
calculated within the 2D framework. 
%This is not surprising since the
%instability is fundamentally driven by viscous and self-gravity forces in the
%momentum equations, not the energy equation.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous, 
    self-gravitating modes. The black curves are obtained
    from the 3D eigenvalue problem with radiative diffusion,
    applicable to optically-thick regions ($R\lesssim 60$AU); 
    while the green curves are obtained from the 2D corresponding
    model, which is applicable to both optically-thick and
    optically-thin regimes.  
    \label{rafikov_growth3d}}
\end{figure}

### 2016-01-17 paper/summary.tex
regular update
reg update

\section{Summary and discussion}
In this paper, we study the implications on the stability of
turbulent, cooling accretion disks if said turbulence is modeled as an
effective, Navier-Stokes viscosity. This viscosity provides a 
background heating and may also be active in the perturbed 
state. We pay particular attention to 
protoplanetary disks subject to `gravito-turbulence'. 
Our framework provide an alternative interpretation of existing
results on PPD fragmentation established from direct numerical
simulations. 
{\bf cooling removes pressure support, viscosity removes rotational support}
{\bf our conclusions do require input of emperical resuls from numerical
sims}

@@

We derive the disperion relation for local axisymmetric density waves 
subject to external heating and cooling (Eq. \ref{inviscid}), but
without viscous effects on the waves. We find
the sufficient condition for stability 
is \emph{independent} of the cooling rate. Long cooling times can
formally lead to instability, but the growth timescales may be
uninterestingly long. We obtain an analytic expression for the
dimensionless cooling time, $\beta_*$, required to remove pressure
support against self-gravity on radial lengthscales comparable to the 
disk scale height. We find $\beta_*$ reproduces the fragmentation
boundary found in previous numerical simulations (Table
\ref{bstar_compare}).

%viscous GI ad connection with frag

We then include viscosity in the linear problem and generalize the
`viscous gravitational instability', previously discussed in 
simpler disk models \citep{lynden-bell74,gammie96,hunter83}, to include
heating/cooling and three-dimensionality. Our generalized framework is
readily applied to realistic protoplanetary disk models. For
gravito-turbulent PPDs, we find viscous GI grows on orbital time-scales
beyond $\sim 100\mathrm{AU}$. This means the assumed gravito-turbulent
steady state should not exist beyond dynamical time-scales in the
outer parts of PPDs.

%viscosity > 0.1, beta < 3 - 4

Our results bear striking similarities with the current consensus on
PPD fragmentation. It is generally accepted that PPDs can fragment
beyond a few tens of $\mathrm{AU}$. This conclusion results from
applying a Toomre stability condition (e.g. $Q\lesssim 1.5$) and a
cooling time or viscosity criteria (e.g. $\beta\lesssim 3$ or
$\alpha\gtrsim 0.1$), the latter of which is
based on numerical simulations. We find these
emperical conditions essentially translates to viscous GI on dynamical
time-scales. This suggests that gravito-turbulent
PPDs fragment because the associated viscous stresses further
destabilize the disk against self-gravity.

%ultimately: why do disks fragment? viscosity may play a role

% also conclude this region of

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling,
turbulent accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We model turbulence as an effective, 
Navier-Stokes viscosity. This viscosity provides a 
background heating to balance the imposed cooling rate,  
and may also be active in the perturbed state. We provide physical
interpretations of disk fragmentation, observed in 
numerical simulations, as a secondary GI from a 
gravito-turbulent state, driven by cooling and/or viscosity. 
%{\bf cooling removes pressure support, viscosity removes rotational support}
%{\bf our conclusions do require input of emperical resuls from numerical
%sims}
%going beyond toomre 
%move to subsection?
While viscosity and cooling are inversely related through thermal
balance, they independently enhance GI: cooling reduces thermal
support; and viscous forces compromise rotational support.

@@

The effect of cooling on GI is quantified by the dispersion
relation Eq. \ref{inviscid}. 
%We derive the disperion relation for local axisymmetric density waves 
%subject to external heating and cooling (Eq. \ref{inviscid}), but
%without viscous effects. 
We find
the sufficient condition for stability is \emph{independent} of the
cooling rate, provided it is non-zero and finite.   
Long cooling times can
formally lead to instability, but the growth timescales may be
uninterestingly long.

%We obtain an analytic expression for the
%dimensionless cooling time, $\beta_*$, required to remove pressure
%support against self-gravity on radial lengthscales comparable to the 
%disk scale height. We find $\beta_*$ reproduces the fragmentation
%boundary found in previous numerical simulations (Table
%\ref{bstar_compare}).

%viscous GI (new: 3D, cooling, ext. heating)

We generalize the `viscous gravitational instability', previously discussed in 
simpler disk models \citep{lynden-bell74,gammie96,hunter83}, to include
heating/cooling and three-dimensionality.

%application to PPD

Our generalized framework is
readily applied to realistic protoplanetary disk models. For
gravito-turbulent PPDs, we find viscous GI grows on orbital time-scales
beyond $\sim 100\mathrm{AU}$. This means the assumed gravito-turbulent
steady state should not exist beyond dynamical time-scales in the
outer parts of PPDs.

Our results bear striking similarities with the current
consensus on PPD fragmentation. It is generally ac-
cepted that PPDs can fragment beyond a few tens of
AU. This conclusion results from applying a Toomre
stability condition (e.g. $Q\lesssim 1.5$) and a cooling time or
viscosity criteria (e.g. $\beta\lesssim3$ or $\alpha\gtrsim 0.1$), the latter of
which is based on numerical simulations. We find these
emperical conditions essentially translates to viscous GI
on dynamical time-scales. This suggests that gravito-
turbulent PPDs fragment because the associated viscous
stresses further destabilize the disk against self-gravity.

%ultimately: why do disks fragment? viscosity may play a role
%details of thermo doesn't matter if it's viscosity that leads to
%fragmentation

\subsection{Relation to numerical simulations}\label{prev_works}

%frag due to cooling -> structures less than H
%frag due to viscosity -> structures more than H
%stochastic frag
%non-convergence with resolution

\subsection{MHD turbulence}
%application to gravito-turbulence but actually linear theory does not assume
%actually application to 
%MHD turbulence prob makes more sense 
%require `scale separation' for GI turbulence (small-scale GI provide
%turbulence, large-scale GI for fragmentation)

\subsection{Outstanding issues}
%non-axisymmtric problem 
%viscosity model (maybe it doesn't behave like a power law)
%chosen a particular mu and lambda
